---
title: "Babette practical"
author: "Thijs Janzen & Rampal Etienne"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(babette)
library(ape)
library(beautier)
library(mauricer)
library(tracerer)
```

## Inferring phylogenies

There are many softwares to infer phylogenies from a sequence. Here, in
this practical, we will make use of the R package 'babette', which
provides an interface to make use of the software BEAST2 (Bayesian
Evolutionary Analysis Sampling Trees). Although BEAST2 by default
provides an interface to setup a phylogenetic analysis, this interface
can be tricky to work with, and it can be hard to replicate analyses
made using this interface. Instead, Richel Bilderbeek has developed the
R package 'babette', which allows us to write an R script to perform a
BEAST2 analysis.

### Installing required packages

To be able to perform our analyses, there are a number of packages that
need to be installed, including of course BEAST2. Let's get started!

```{r installation, eval = FALSE}
install.packages("ape")
install.packages("beastier")
install.packages("tracerer")
install.packages("mauricer")
install.packages("beautier")

# this installs beast2:
remotes::install_github("richelbilderbeek/beastierinstall")
beastierinstall::install_beast2()

install.packages("babette")

# and for plotting:
install.packages("tidyverse")
```

Now, we have all packages installed. Unfortunately, BEAST2 needs Java to
run. This can turn out to be quite finnicky to get to work for some
laptops. Please first try:

```{r install java, eval = FALSE}
install.packages("rJava")
```

If that does not work to install rJava, please also see these notes:
<https://github.com/ropensci/babette/blob/master/doc/rjava.md>

Please note that getting rJava and Java to run can, in some cases (e.g. with
older laptops), be a complicated process. If you don't manage to get Java to run,
please don't spend hours trying to get this to work, but instead see if you can
join forces with a neighbor or fellow-student that did manage to get everything
correctly installed.

All packages can now be loaded:
```{r load_packages, eval = FALSE}
library(tidyverse)
library(babette)
library(ape)
library(beautier)
library(mauricer)
library(tracerer)
```

### Loading data

We can load data into R using the following functions, where we load an
example file from the BEAST2 examples (alternatively, download the file
from Brightspace). We then use the package ape to read the example file.

```{r loading data}
nexus_filename <- beastier::get_beast2_example_filename("dna.nex")

nexus_data <- ape::read.nexus.data(nexus_filename)
output_data <- ape::as.DNAbin(nexus_data)

ape::write.FASTA(output_data, file = "/Users/thijsjanzen/dna.fas")

str(nexus_data)
```

The last command shows us that the dataset we have obtained containes
sequences of 10 quite different species. Our goal in this tutorial is to
infer their relationships and find when they diverged from each other.

### Setting up babette analysis

To set up a babette analysis, we have to specify a number of
prerequisites \#### Tree prior To set up the tree prior, e.g. we
predefine a model of how we expect species to branch over time, we can
make use of the predefined tree priors in beautier, such as: Yule,
Birth-Death and different coalescent priors (see
?beautier::create_tree_priors for an overview of all available priors).\
Here, we will make use of a birth-death tree prior.

```{r tree_prior}
bd_prior <- beautier::create_tree_prior_bd()
```

#### Clock model

The clock model translates genetic differences into time units. In other
words, it determines how 'fast' the clock ticks. However, typically we
find that the clock does not tick equally fast along all branches, and
this needs to be corrected for. Instead, typically it is assumed that
the clock rates of each branch are slightly different, but all belong to
the same distribution. Babette currently supports two clock models: the
strict clock model, that assumes clock rates along all branches are
identical, and the relaxed log normal clock model, that assumes clock
rates along all branches are drawn from a log normal distribution. We
will use the latter.

```{r clock_model}
rln_clock <- beautier::create_clock_model_rln()
```


#### Substitution model

The substition model chooses how bases mutate into each other. In babette, there
are a number of substitution models available, such as the JC69, HKY, TN93 and
GTR model. The simplest model is the JC69 model; the JC69 model assumes that
all bases have equal probability of mutating into each other, e.g. a mutation
from A to T has the same probability as a mutation from A to G. On the other end
of the extreme, The Generalised time reversible (GTR) model fits all 6 different
substitution rates for each possible substitution (please verify that you can
capture all possible substitutions between 4 bases, with 6 rates). The other
available models pose some restrictions on the potential rates, and fall in
between the JC69 and GTR model in terms of complexity.
In this tutorial we will use the JC69 model for simplicity.

```{r sub_model}
sub_model <- beautier::create_site_model_jc69()
```

#### Define MCMC

BEAST2 makes use of Markov Chain Monte Carlo (MCMC) to find the best
fitting parameters. MCMC is a numerical method aimed at stochastically
(the Monte Carlo part) exploring parameter space, weighed by the fit. In
other words: we aim to explore most the area with the best fit, and not
spend time in parameter space with poor fit. Ideally, we would map the
entirety of parameter space to find the perfect fit, however given the
number of parameters this is impossible, and instead we will make use of
MCMC. The core idea is that we pick a parameter combination and
calculate the likelihood, then, we slightly perturb the parameter
combination and calculate the likelihood again. Depending on the
improvement of the likelihood, we move to the new parameter combination
or not. Iteratively doing so creates a parameter 'chain', where each
entry only depends on the entry before (e.g. a Markov Chain). Thus
moving through parameter space is an efficient way of exploring only
those areas with high likelihood. However, values are often
auto-correlated, and to avoid this, a common practice is to only store
every Nth parameter combination. Here, we will use an interval of 1000,
and generate a chain of 1e6 combinations:

```{r chain_setup}
mcmc_settings <- beautier::create_mcmc(chain_length = 1e6, 
                                       store_every = 1000)
```

## Running Babette

Now we can combine our priors and mcmc settings into an inference model,
and pass this inference model to babette to start the inference

```{r inf_model}
inf_model <- beautier::create_inference_model(tree_prior = bd_prior,
                                              site_model = sub_model,
                                              clock_model = rln_clock,
                                              mcmc = mcmc_settings)

beast_result <- babette::bbt_run_from_model(fasta_filename = "dna.fas",
                                            inference_model = inf_model,
                                            beast2_options = beastier::create_beast2_options(rng_seed = 42))
```

### Analysing results

Now that we have our results, we can plot them and see how we did:

```{r plot_things}
beast_result$estimates %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free")
```

In these plots, we see on the x-axis our sample (these are one thousand
points, right?), and on the y-axis the associated value. Shown are
results for 12 different optimized parameters. If our Markov Chain
converged well, we should see 'Hairy Caterpillars' for all parameters,
indicating that the chain was circling around the optimum and sampling
well. Looking at our results, we immediately see that a number of plots
show very different results in the beginning, these are the result of
initialization of the chain. This is a common occurance, and is
typically removed from the dataset as 'burn in'. Here, we will remove
the first 10% of the samples to get rid of these poor datapoints.
Typically, removal of 10-20% of the data is considered acceptable.

```{r remove burnin}
beast_result$estimates <- tracerer::remove_burn_ins(beast_result$estimates, 0.1)
beast_result$estimates %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free")
```

Upon re-plotting the data, we see that now, all variables have turned
into hairy caterpillars. A further check of whether our inference was
long enough is by calculating the Effective Sample Size (ESS), this
should be at least 200 for each variable:

```{r ess}
tracerer::calc_esses(beast_result$estimates, sample_interval = 1000)
```

For some variables (TreeHeight and rate.variance) values are not 200
yet. The best way forward would be to either repeat the analysis with a
longer chain, or continue the existing chain. For the purpose of this
tutorial, we will continue with the existing results, also considering
values are already quite close to 200.

The MCMC results have also generated 1000 trees. Of these, again we have
to remove the first 10%, and then we can plot these trees to see whether
they have converged unto a unifying tree.

```{r trees}
all_trees <- beast_result$dna_trees
start <- floor(0.1 * length(all_trees))
end <- length(all_trees)
all_trees <- all_trees[start:end]
```

Having removed the burnin, we can now plot our results in two ways: 1)
using a densi-tree, which superimposes all trees on top of each other
and 2) using a consensus tree, which calculates the 'average' resulting
tree from our distribution of 900 trees.

```{r plotting}
babette::plot_densitree(all_trees, alpha = 0.1,
                        type = "cladogram")

cons_tree <- phytools::consensus.edges(trees = all_trees)
plot(cons_tree)
```

In the consensus tree, we see a topology that seems to make sense -
species that are close related appear close to each other. However, the
consensus tree does not tell us anything about uncertainty in the tree
estimation process. The densitree does. And looking at the densitree, we
see some interesting patters, mainly regarding the root, where some
trees in the posterior have swapped the position of humans and chicken.

### Substitution model

How does the substitution model affect our results? We can easily repeat the
analysis, but now change the substitution model. Let's do so!

```{r GTR}
sub_model <- beautier::create_site_model_gtr()

inf_model <- beautier::create_inference_model(tree_prior = bd_prior,
                                              site_model = sub_model,
                                              clock_model = rln_clock,
                                              mcmc = mcmc_settings)

beast_result_gtr <- babette::bbt_run_from_model(fasta_filename = "dna.fas",
                                            inference_model = inf_model,
                                            beast2_options = beastier::create_beast2_options(rng_seed = 4321))
```

If we plot our results, we will see that additional variables have been added
to reflect estimates for the different substitution rates:

```{r plot_gtr}
beast_result_gtr$estimates %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free")
```

This is not yet very informative due to the burnin effects. Also, perhaps it is
interesting to look at the median estimates for the different rates:

```{r gtr_burnin plot}
gtr_estimates <- tracerer::remove_burn_ins(beast_result_gtr$estimates, 0.1)
gtr_estimates %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free")

gtr_estimates %>%
  gather(key = "statistic", val = "value", c(rateAC, rateAG, rateAT, rateCG, rateGT)) %>%
  group_by(statistic) %>%
  summarise("median" = median(value))
```

From these values, it is clear that substitutions from A to C (0.6) and A to G
(0.75) where estimated to be most frequent, with substitutions from A to T (0.41),
C to G (0.13) and G to T (0.09) much less frequent.
For the JC69 model we had no rate estimates, as there all substitutions are
equally likely (so the rate is '1').

Does a different substitution model affect our findings? We collect
again the trees after removal of burnin, and then plot the tree
side by side with the JC69 findings. Also, we add a plot where we
plot the lineages over time to see if our substitution model has
changed anything.
```{r gtr trees}
gtr_trees <- beast_result_gtr$dna_trees
start <- floor(0.1 * length(gtr_trees))
end <- length(gtr_trees)
gtr_trees <- gtr_trees[start:end]
babette::plot_densitree(gtr_trees, alpha = 0.1, type = "cladogram")

cons_tree_gtr <- phytools::consensus.edges(trees = all_trees)
par(mfrow = c(1, 2))
plot(cons_tree, main = "JC69")
plot(cons_tree_gtr, main = "GTR")

ape::ltt.plot(cons_tree)
ape::ltt.lines(cons_tree_gtr, col = "blue")
```

It seems the data is so strongly divergent, that it overwhelms
choices about the substitution model.

### Root age

It is clear from the densitree, that the root age has been inferred
incorrectly: the root is now placed at 0.25 MY, which seems very, very
young. Indeed, birds and mammals are thought to have diverged from each
other around 319 MYA (sd = 1.64) (timetree.org). We can add this
information into our analysis by putting a prior on the crown age.

```{r mrca_prior}
mrca_prior <- beautier::create_mrca_prior(
     taxa_names = beautier::get_taxa_names("dna.fas"),
     is_monophyletic = TRUE,
     mrca_distr = beautier::create_normal_distr(mean = 319, sigma = 1.64))
```

Now we have to re-specify our inference model to include the mrca prior,
and because we are using an mrca prior, we also have to force usage of
beauti 2.6. Lastly, because we are adding complexity to the inference, we expect
the chain to take longer to converge. To anticipate this, we have chosen a longer
chain length.

```{r mrca_inference_model}
mcmc_settings <- beautier::create_mcmc(chain_length = 1e6, store_every = 1000)


beauti_options <- beautier::create_beauti_options_v2_6()

inf_model <- beautier::create_inference_model(tree_prior = bd_prior,
                                              clock_model = rln_clock,
                                              site_model = sub_model,
                                              mcmc = mcmc_settings,
                                              mrca_prior = mrca_prior,
                                              beauti_options = beauti_options)
```

And now we can repeat the analysis:

```{r babette_with_mrca}
beast_result <- babette::bbt_run_from_model(fasta_filename = "dna.fas",
                                            inference_model = inf_model,
                                            beast2_options = beastier::create_beast2_options(rng_seed = 42))
```

Again, we first plot our MCMC chains:

```{r plot_mrca}
beast_result$estimates %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free")
```

Clearly, we have strong initialization effects. Let's remove 20% this
time:

```{r burnin_mrca}
estimates <- tracerer::remove_burn_ins(beast_result$estimates, 0.4)
estimates %>%
  gather(key = "statistic", val = "value", -c(Sample)) %>%
  ggplot(aes(x = Sample, y = value)) +
  geom_line() +
  facet_wrap(~statistic, scales = "free")
tracerer::calc_esses(estimates, sample_interval = 1000)
```

Having lengthened the chain now pays off, as almost all ESS values are > 200.

We again remove the faulty trees as well, and plot the distribution of trees:
```{r burnin_mrca_trees}
all_trees <- beast_result$dna_trees
start <- floor(0.4 * length(all_trees))
end <- length(all_trees)
all_trees <- all_trees[start:end]

babette::plot_densitree(all_trees, alpha = 0.1,
                        type = "cladogram")
```

Now, we see that the crown of the tree has moved towards 320MY, and that
the other branching points have moved accordingly. Our tree looks much
better now!

## Outlook

We have made a number of simplifying assumptions, and we have also left
some settings at their default. For instance, we have, in this tutorial,
not looked into setting a prior on the substitution model. To have a
look at other potential priors you can set and options you can explore,
please have a look at the help page for the function
'beautier::create_inference_model' for further information.

# Exercise
Now that you've got a feel for the analysis pipeline, it is your turn! Please
load DNA sequences for 12 primate species, and reconstruct their phylogeny. Can
you also infer the age of the most recent common ancestor of Humans and
Chimpansee?
If you have time: how does the substition model influence your findings? Why /
Why not?

You can load the DNA sequences from the beast2 examples as follows:

```{r loading primates}
nexus_filename <- beastier::get_beast2_example_filename("Primates.nex")
fasta_filename <- "primates.fas"

nexus_data <- ape::read.nexus.data(nexus_filename)
output_data <- ape::as.DNAbin(nexus_data)

ape::write.FASTA(output_data, file = fasta_filename)
str(nexus_data)
```

Once again: if you can not get rJava to work, please do not spend hours to fix
this, but rather see if your neighbor has been able to install rJava. Then, work
together to finish this exercise.

Good luck!
